---
title: "exploratory_data_analysis"
author: "Jonny Nelson"
date: "07/02/2022"
output: html_document
---

# LendingClub: Predicting Loan Defaults
* who is likely to default?
* who to lend to in the future?

```{r}
library(janitor)
library(tidyverse)

loans_dictionary <- read_csv("defaults_data/LCDataDictionary.csv")
loans <- read_csv("defaults_data/lending_club_loans.csv")
knitr::kable(loans_dictionary)
```

<br>

::: {.emphasis}
## Noteable points
* acc_open_past_24mths - What are number of trades? - All NA values?
* addrs_state - Mutate on the state name for geospatial mapping
* annual_inc_joint & application_type - can be selected out as all registrations provided are application_type == INDIVIDUAL - 3 NA's for application type.
* avg_cur_bal, bc_open_to_buy, bc_util - All NA's
* chargeoff_within_12_mths - only 148 NA's - otherwise == 0
* collection_recovery_fee - LendingClub expense
* collections_12_mths_ex_med - No of collections
* delinq_2yrs - 30+ past due incidence rate
* delinq_amnt - loads of NA's and 0's
* desc - only avaliable for 1/4 of observations - text analysis???
* dti - debt obligations/monthly income
* dti_joint - NA's as expected
* earliest_cr_line - credint line opening month
* emp_length - employment length time
* emp_title - Employer Title replaces Employer Name for all loans listed after 9/23/2013 - investigate further
* fico_range -> get mean?
* funded_amnt - $ commited to that loan at that point in time
* funded_amnt_inv - $ funded by investors? comming from peer-to-peers lending?
* grade - need to figure out different grades of loans
* home_ownership - rent, own, mortgage or other
* id - unique for loan
* il_util - ratio of total current balance to credit limit on all accounts? Look into this one further
* initial_list_status - W and F? 
* inq_last_6mths - no of inquiries in previous 6 months
* inq_fi and inq_last_12m - All NA's
* installment - monthly's
* int_rate - interest rate
* issue_d - issue date
* last_credit_pull_d - most recent pulled credit for loan
* last_fico's - last pulled fico's - what is the last pulled?
* loan_amnt 
* loan_status - find out difference between observations . . .
* max_bal_bc - All NA's
* member_id - unique id for the member - all values are distinct
* mo_sin . . . - all NA's
* mort_acc - 
:::

```{r}
knitr::kable(loans_dictionary)
```


```{r}
loans %>%
  select(mo_sin_old_il_acct, mo_sin_old_rev_tl_op, mo_sin_rcnt_rev_tl_op, mo_sin_rcnt_tl) 

sum(is.na(loans$mo_sin_rcnt_tl))

range(loans$funded_amnt)

unique(loans$loan_status)

n_distinct(loans$member_id)

# summary(loans)
```

