---
title: "ML"
author: "Jonny Nelson"
date: "14/02/2022"
output: html_document
---

# Reading in Libraries and Data

```{r}
library(janitor)
library(tidyverse)
library(here)
library(forcats)

loans_clean <- read_csv(here("clean_data/loans_clean.csv"))
```

# Variable Engineering

```{r}
lc_loans <- loans_clean %>%
  filter(loan_status %in% c("Charged Off", "Fully Paid")) %>%
  mutate(emp_length = factor(x = emp_length,
                             levels = c("< 1 year", "2 years", "3 years", "4 years", "5 years", "6 years", "7 years", "8 years", "9 years", "10+ years"),
                             labels = c("More than 1" , "2", "3", "4", "5", "6", "7", "8", "9", "10 +")),
         desc = if_else(desc %in% NA, 0, 1 )) %>%
  mutate(emp_length = fct_explicit_na(emp_length, "unknown"))

# Write the cleaned, ML ready into the clean data folder
# write_csv(x = lc_loans, here("clean_data/lc_loans.csv"))

unique(lc_loans$emp_length)
```

# Variable Selection

```{r}
ml_loans <- lc_loans %>%
  select(c(loan_status, loan_amnt, term, int_rate, emp_length, home_ownership, installment, annual_inc, verification_status, pymnt_plan, desc, purpose, debt_to_income_ratio, derog_pub_rec, mean_fico_scores, grade))
head(ml_loans)

## Retro-actively taking out issue_d and sub_grade as these are variables with very high carnality. May put back into model at a later date

## Retro-actively taking out revol_util_rate as is has NA values, messing up the backwards selection. No clear relevance to loan_status

# Installment and payment plan is given after the loan has been approved
```

### Writing the Data for Python ML

```{r}
# write_csv(x = ml_loans_converted, here("clean_data/ml_data.csv"))
```

# 1. Examine relationships between Response and Predictor variables in ggpairs

## Splitting the data for ggpairs

```{r}
split_1 <- ml_loans %>%
  select(c(loan_status, loan_amnt, term, int_rate))
split_2 <- ml_loans %>%
  select(c(loan_status, installment, emp_length))
split_3 <- ml_loans %>%
  select(c(loan_status, home_ownership, annual_inc, verification_status))
split_4 <- ml_loans %>%
  select(c(loan_status, pymnt_plan, desc))
split_5 <- ml_loans %>%
  select(c(loan_status, purpose, debt_to_income_ratio, derog_pub_rec))
split_6 <- ml_loans %>%
  select(c(loan_status, revol_util_rate, mean_fico_scores, grade))
```

## Visualising using ggpairs

```{r}
library(GGally)
```

## Split 1

```{r}
split_1 %>%
  ggpairs()
```
* The distributions for the interest rates are slightly different based on loan_status. Loans with higher interest rates have slightly higher rates of defaults.
* Far more loans default with term lengths of 60 months compared to loans of 36 months.

### Investigating Interest Rates:

```{r}
ml_loans %>%
  ggplot() +
  aes(x = loan_status, y = int_rate) +
  geom_boxplot()
```

### Investigating Term Length:

```{r}
ml_loans %>%
  ggplot() +
  geom_bar(aes(x = loan_status, fill = term), position = "dodge") + 
  scale_fill_discrete() + 
  theme_bw() +
  theme(axis.line = element_line(colour = "black"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    panel.background = element_blank()) +
  xlab("Loan Status") +
  ylab("Count") +
  ggtitle("Count for Loan Status by Term Length") +
  guides(fill = guide_legend(title = "Term Length")) 
```

## Split 2

```{r}
split_2 %>%
  select(-sub_grade) %>%
  ggpairs()

# sub_grade has a very high cardinality, makes many dummy variables so is too difficult to run on ggpairs, may be able to make into a continuous variable.

# unique(ml_loans$sub_grade)

# 35 specific sub_grades could be made into a continuous variables from 0-35. 
```

* No obvious correlations here. Employment length may hold some useful information in one of the specific categories.


## Split 3

```{r}
split_3 %>%
  ggpairs()
```

* Hard to draw anything meaning form these variables.

### Annual Income 

```{r}
ggplot(ml_loans) +
  geom_jitter(aes(x = annual_inc, y = loan_status), shape = 1, 
              position = position_jitter(h = 0.5)) +
  xlim(0, 500000)

# Closer inspection of annual income shows people with very large annual incomes are far more likely to fully pay back a loan. 
```

## Split 4

```{r}
split_4 %>%
  select(-issue_d) %>%
  ggpairs()

# Issue date taken away due to high carnality. May warrant further exploration in the extension.
```

* Again. No major correlations identified here.

## Split 5

```{r}
split_5 %>%
  ggpairs()
```

* No major correlations here

## Split 6

```{r}
split_6 %>%
  ggpairs()
```

* All variables show some correlation with loan_status

### Visualising the Mean FICO scores

```{r}
ggplot(ml_loans) +
  geom_jitter(aes(x = mean_fico_scores, y = loan_status), shape = 1, 
              position = position_jitter(h = 0.25))

# Divergence between fully paid and charged off begins around a mean FICO score of 750
```


# 2. Convert all character to factor columns

```{r}
ml_loans_converted <- ml_loans %>%
  mutate(loan_status = if_else(loan_status == "Fully Paid", 0, 1)) %>%
  mutate_if(is_character, as_factor) 
```

# 3. Investigate the forwards and backwards selection

## Backwards Selection

```{r}
# Backwards Selection to Identify the most notable variables

library(leaps)

regsubsets_backward <- regsubsets(loan_status ~ ., data = ml_loans_converted, nvmax = 20, method = "backward")

plot(regsubsets_backward, scale = "bic")

# Looks like term_60_months is first, then installments, purposeCAR and then home_ownershipNONE
```

```{r}
summary_regsubsets_backward <- summary(regsubsets_backward)
summary_regsubsets_backward

# int_rate, term60, purposesmall_business in order of importance
```

## Forward Selection

```{r}
# Backwards Selection to Identify the most notable variables

library(leaps)

regsubsets_forward <- regsubsets(loan_status ~ ., data = ml_loans_converted, nvmax = 20, method = "forward")

plot(regsubsets_forward, scale = "bic")


# Looks like term_60_months is first, then installments, purposeCAR and then home_ownershipNONE.
```

```{r}
summary_regsubsets_forward <- summary(regsubsets_forward)
summary_regsubsets_forward

# int_rate, term60, purposesmall_business in order of importance
```

**Forwards and Backwards Selection give the same result**

```{r}
plot(summary_regsubsets_forward$rsq, type = "b")
```

```{r}
plot(summary_regsubsets_forward$bic, type = "b")
```




# 4. Perform a Logistic Regression

* # Looks like term_60_months is first, then installments, purposeCAR and then home_ownershipNONE

```{r}
library(fastDummies)


```


```{r}
# Term Length
model_1_pred_term <- glm(loan_status ~ term,
                         data = ml_loans_converted,
                         family = binomial(link = "logit"))

# Installments 
model_2_pred_installments <- glm(loan_status ~ installment,
                                 data = ml_loans_converted,
                                 family = binomial(link = "logit"))

# Purpose == Car 
model_3_pred_purposecar <- glm(loan_status ~ purpose,
                               data = ml_loans_converted,
                               family = binomial(link = "logit"))

# Home Ownership == None
model_4_pred_homeownershipnone <- glm(loan_status ~ home_ownership,
                                      data = ml_loans_converted,
                                      family = binomial(link = "logit"))

colnames(ml_loans_converted)
```

