---
title: "ML"
author: "Jonny Nelson"
date: "14/02/2022"
output: html_document
---

# Reading in Libraries and Data

```{r}
library(janitor)
library(tidyverse)
library(here)
library(forcats)

loans_clean <- read_csv(here("clean_data/loans_clean.csv"))
```

# Variable Engineering

```{r}
lc_loans <- loans_clean %>%
  filter(loan_status %in% c("Charged Off", "Fully Paid")) %>%
  mutate(emp_length = factor(x = emp_length,
                             levels = c("< 1 year", "2 years", "3 years", "4 years", "5 years", "6 years", "7 years", "8 years", "9 years", "10+ years"),
                             labels = c("More than 1" , "2", "3", "4", "5", "6", "7", "8", "9", "10 +")),
         desc = if_else(desc %in% NA, 0, 1 )) %>%
  mutate(emp_length = fct_explicit_na(emp_length, "unknown"))

# Write the cleaned, ML ready into the clean data folder
# write_csv(x = lc_loans, here("clean_data/lc_loans.csv"))

unique(lc_loans$emp_length)
```

# Variable Selection

```{r}
ml_loans <- lc_loans %>%
  select(c(loan_status, loan_amnt, term, int_rate, emp_length, home_ownership, installment, annual_inc, verification_status, pymnt_plan, desc, purpose, debt_to_income_ratio, derog_pub_rec, mean_fico_scores, grade))
head(ml_loans)

## Retro-actively taking out issue_d and sub_grade as these are variables with very high carnality. May put back into model at a later date

## Retro-actively taking out revol_util_rate as is has NA values, messing up the backwards selection. No clear relevance to loan_status

# Installment and payment plan is given after the loan has been approved
```

### Writing the Data for Python ML

```{r}
# write_csv(x = ml_loans_converted, here("clean_data/ml_data.csv"))
```

# 1. Examine relationships between Response and Predictor variables in ggpairs

## Splitting the data for ggpairs

```{r}
split_1 <- ml_loans %>%
  select(c(loan_status, loan_amnt, term, int_rate))
split_2 <- ml_loans %>%
  select(c(loan_status, installment, emp_length))
split_3 <- ml_loans %>%
  select(c(loan_status, home_ownership, annual_inc, verification_status))
split_4 <- ml_loans %>%
  select(c(loan_status, pymnt_plan, desc))
split_5 <- ml_loans %>%
  select(c(loan_status, purpose, debt_to_income_ratio, derog_pub_rec))
split_6 <- ml_loans %>%
  select(c(loan_status, mean_fico_scores, grade))
```

## Visualising using ggpairs

```{r}
library(GGally)
```

## Split 1

```{r}
split_1 %>%
  ggpairs()
```
* The distributions for the interest rates are slightly different based on loan_status. Loans with higher interest rates have slightly higher rates of defaults.
* Far more loans default with term lengths of 60 months compared to loans of 36 months.

### Investigating Interest Rates:

```{r}
ml_loans %>%
  ggplot() +
  aes(x = loan_status, y = int_rate) +
  geom_boxplot()
```

### Investigating Term Length:

```{r}
ml_loans %>%
  ggplot() +
  geom_bar(aes(x = loan_status, fill = term), position = "dodge") + 
  scale_fill_discrete() + 
  theme_bw() +
  theme(axis.line = element_line(colour = "black"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    panel.background = element_blank()) +
  xlab("Loan Status") +
  ylab("Count") +
  ggtitle("Count for Loan Status by Term Length") +
  guides(fill = guide_legend(title = "Term Length")) 
```

## Split 2

```{r}
split_2 %>%
  select(-sub_grade) %>%
  ggpairs()

# sub_grade has a very high cardinality, makes many dummy variables so is too difficult to run on ggpairs, may be able to make into a continuous variable.

# unique(ml_loans$sub_grade)

# 35 specific sub_grades could be made into a continuous variables from 0-35. 
```

* No obvious correlations here. Employment length may hold some useful information in one of the specific categories.


## Split 3

```{r}
split_3 %>%
  ggpairs()
```

* Hard to draw anything meaning form these variables.

### Annual Income 

```{r}
ggplot(ml_loans) +
  geom_jitter(aes(x = annual_inc, y = loan_status), shape = 1, 
              position = position_jitter(h = 0.5)) +
  xlim(0, 500000)

# Closer inspection of annual income shows people with very large annual incomes are far more likely to fully pay back a loan. 
```

## Split 4

```{r}
split_4 %>%
  select(-issue_d) %>%
  ggpairs()

# Issue date taken away due to high carnality. May warrant further exploration in the extension.
```

* Again. No major correlations identified here.

## Split 5

```{r}
split_5 %>%
  ggpairs()
```

* No major correlations here

## Split 6

```{r}
split_6 %>%
  ggpairs()
```

* All variables show some correlation with loan_status

### Visualising the Mean FICO scores

```{r}
ggplot(ml_loans) +
  geom_jitter(aes(x = mean_fico_scores, y = loan_status), shape = 1, 
              position = position_jitter(h = 0.25))

# Divergence between fully paid and charged off begins around a mean FICO score of 750
```


# 2. Convert all character to factor columns

```{r}
ml_loans_converted <- ml_loans %>%
  mutate(loan_status = if_else(loan_status == "Fully Paid", 0, 1)) %>%
  mutate_if(is_character, as_factor) 
```

# 3. Investigate the forwards and backwards selection

## Backwards Selection

```{r}
# Backwards Selection to Identify the most notable variables

library(leaps)

regsubsets_backward <- regsubsets(loan_status ~ ., data = ml_loans_converted, nvmax = 20, method = "backward")

plot(regsubsets_backward, scale = "bic")

# Looks like term_60_months is first, then installments, purposeCAR and then home_ownershipNONE
```

```{r}
summary_regsubsets_backward <- summary(regsubsets_backward)
summary_regsubsets_backward

# int_rate, term60, purposesmall_business in order of importance
```

## Forward Selection

```{r}
# Backwards Selection to Identify the most notable variables

library(leaps)

regsubsets_forward <- regsubsets(loan_status ~ ., data = ml_loans_converted, nvmax = 20, method = "forward")

plot(regsubsets_forward, scale = "bic")


# Looks like term_60_months is first, then installments, purposeCAR and then home_ownershipNONE.
```

```{r}
summary_regsubsets_forward <- summary(regsubsets_forward)
summary_regsubsets_forward

# int_rate, term60, purposesmall_business in order of importance
```

**Forwards and Backwards Selection give the same result**

```{r}
plot(summary_regsubsets_forward$rsq, type = "b")
```

```{r}
plot(summary_regsubsets_forward$bic, type = "b")
```




# 4. Perform a Logistic Regression

## Checking for Aliased variables

* # Looks like term_60_months is first, then installments, purposeCAR and then home_ownershipNONE

```{r}
alias(loan_status ~ ., data = ml_loans_converted)

# No aliased variables found
```

## Splitting the data into Test Train Sets

```{r}
set.seed(7)

test_indices <- sample(1:nrow(ml_loans_converted), size = as.integer(nrow(ml_loans_converted) * 0.2))

train <- ml_loans_converted %>%
  slice(-test_indices)

test <- ml_loans_converted %>%
  slice(test_indices)

# sanity check
nrow(train) + nrow(test) == nrow(ml_loans_converted)
```

## Checking the distributions

```{r}
train %>%
  tabyl(loan_status)
```

```{r}
test %>%
  tabyl(loan_status)
```

* Loan status is nicely split in the train and test split. 

# Model Building

```{r}
# Loading the library
library(glmulti)

# Model 1
model_1 <- glm(loan_status ~ ., data = train, family = binomial(link = "logit"))

summary(model_1)
```

## Best Model with No Interactions

```{r}
glmulti_search_all_mains <- glmulti(
  loan_status ~ ., 
  data = train,
  level = 1,               # No interaction considered
  method = "h",            # Exhaustive approach
  crit = "bic",            # BIC as criteria
  confsetsize = 10,        # Keep 10 best models
  plotty = F, 
  report = T,              # No plots, but provide interim reports
  fitfunction = "glm",     # glm function
  family = binomial(link = "logit")) # binomial family for logistic regression

summary(glmulti_search_all_mains)

 # After 1350 models:
 # Best model: loan_status ~ 1 + term + int_rate + annual_inc + derog_pub_rec
# Crit= 24359.344400846
# Mean crit= 24369.0183010957
```


## Best Model with 2 Pair Interactions 

```{r}
glmulti_search_all_mains_two_pairs <- glmulti(
  loan_status ~ loan_amnt + int_rate + installment + annual_inc + derog_pub_rec, 
  data = train,
  level = 2,               # Interactions considered
  method = "h",            # Exhaustive approach
  crit = "bic",            # BIC as criteria
  confsetsize = 10,        # Keep 10 best models
  marginality = TRUE,      # consider pairs only if both main effects in model
  minsize = 7,             # minsize, maxsize and marginality here force 
  maxsize = 7,             # inclusion of a single pair beyond the five main effects
  plotty = F, 
  report = T,              # No plots, but provide interim reports
  fitfunction = "glm",     # glm function
  family = binomial(link = "logit")) # binomial family for logistic regression

summary(glmulti_search_all_mains_two_pairs)

# After 150 models:
# Best model: loan_status ~ 1 + loan_amnt + int_rate + installment + annual_inc + derog_pub_rec + annual_inc:installment + derog_pub_rec:installment
# Crit= 24428.8534868457
# Mean crit= 24431.9232538522
# Completed.
```

## Best Model considering AIC

```{r}
glmulti_ga_search_with_pairs_aic <- glmulti(
  loan_status ~ ., 
  data = train,
  level = 2,               # Interactions considered
  method = "g",            # Genetic algorithm approach
  crit = "aic",            # AIC as criteria
  confsetsize = 10,        # Keep 10 best models
  marginality = TRUE,      # Consider pairs only if both main effects in model
  plotty = F, 
  report = T,              # No plots, but provide interim reports
  fitfunction = "glm",     # glm function
  family = binomial(link = "logit")) # Binomial family for logistic regression

summary(glmulti_ga_search_with_pairs_aic)
```

## Comparing AUC value

```{r}
library(pROC)
library(modelr)
```

## Model with main BIC

### AUC Train

```{r}
model_with_mains_bic <- glm(
  loan_status ~ term + int_rate + annual_inc + derog_pub_rec,
  data = train,
  family = binomial(link = "logit")
)
roc <- train %>%
  add_predictions(model_with_mains_bic, type = "response") %>%
  roc(response = loan_status, predictor = pred)

auc(roc)
# AUC = 0.682
```

### AUC Test

```{r}
roc <- test %>%
  add_predictions(model_with_mains_bic, type = "response") %>%
  roc(response = loan_status, predictor = pred)

auc(roc)
# AUC = 0.7011
```




## Model with 2 Interactions BIC

### AUC Train

```{r}
model_with_one_pair_bic <- glm(
  loan_status ~ loan_amnt + int_rate + installment + annual_inc + derog_pub_rec + annual_inc:installment,
  data = train,
  family = binomial(link = "logit")
)
roc <- train %>%
  add_predictions(model_with_one_pair_bic, type = "response") %>%
  roc(response = loan_status, predictor = pred)

auc(roc)
# AUC = 0.6793
```

### AUC Test

```{r}
roc <- test %>%
  add_predictions(model_with_one_pair_bic, type = "response") %>%
  roc(response = loan_status, predictor = pred)

auc(roc)
# AUC = 0.7015
```

## Model with 3 Interactions BIC

### AUC Train

```{r}
model_with_two_pair_bic <- glm(
  loan_status ~ loan_amnt + int_rate + installment + annual_inc + derog_pub_rec + annual_inc:installment + derog_pub_rec:installment,
  data = train,
  family = binomial(link = "logit")
)
roc <- train %>%
  add_predictions(model_with_two_pair_bic, type = "response") %>%
  roc(response = loan_status, predictor = pred)

auc(roc)
# AUC = 0.6796
```

### AUC Test

```{r}
roc <- test %>%
  add_predictions(model_with_two_pair_bic, type = "response") %>%
  roc(response = loan_status, predictor = pred)

auc(roc)
# AUC = 0.7011
```

# Final Notes

* Model that is the best is as follows

```{r}
best_model <- glm(
  loan_status ~ loan_amnt + int_rate + installment + annual_inc + derog_pub_rec + annual_inc:installment,
  data = train,
  family = binomial(link = "logit")
)

# Highest AUC value for the test set - 0.7015
```
## Results

```{r}
library(caret)

## Train Control

train_control <- trainControl(method = "repeatedcv", 
                              number = 5,
                              repeats = 100,
                              savePredictions = TRUE, 
                              classProbs = TRUE, 
                              summaryFunction = twoClassSummary)

## Best Model

best_model_cv <- train(best_model$formula,
                       data = ml_loans_converted,
                       trControl = train_control,
                       method = "glm",
                       family = binomial(link = 'logit'))
```

## Up-Sampling

```{r}
# Loading the caret package
library(caret)

# Response as a factor
train_subsampling <- train %>%
  mutate(loan_status = factor(loan_status))


up_train <- upSample(x = train_subsampling[, -ncol(train_subsampling)],
                     y = train_subsampling$loan_status) 

table(up_train$loan_status) 
```

